A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-03 ISMB/introduction_ISMB.tex
typos

In this paper, we introduce \ourprog, a RNA secondary structure design algorithm that benefits of our recent algorithmic advances \cite{Reinharz:2013aa} to expand our original \RNAensign algorithm \cite{Levin:2012kx}. \ourprog addresses previous limitations of \RNAensign and offers new functionalities. First, while our previous program had a running time complexity of $\mathcal{O}(n^5)$, \ourprog now runs in linear-time and space complexity, allowing it to demonstrate similar speeds as any local search algorithm. Next, \ourprog is \textit{seed-less}. Unlike \RNAensign, it does not require a seed sequence to initiate its search. Finally, \ourprog implements a novel algorithm using weighted sampling techniques~\cite{Bodini2010} that enables us to control, for the first time, \textit{explicitly} the \GCContent of the solution. This functionality is essential because wild sequences within living organisms often present medium or low \GCContent, presumably to offer better transcription rates and/or structural plasticity. Previous programs do not allow to control this parameter and tend to output sequences with high \GCContent{}s.

>>>>

In this paper, we introduce \ourprog, a RNA secondary structure design algorithm that benefits of our recent algorithmic advances \cite{Reinharz:2013aa} to expand our original \RNAensign algorithm \cite{Levin:2012kx}. \ourprog addresses previous limitations of \RNAensign and offers new functionalities. First, while our previous program had a running time complexity of $\mathcal{O}(n^5)$, \ourprog now runs in linear-time and space complexity, allowing it to demonstrate similar speeds as any local search algorithm. Next, \ourprog is \textit{seed-less}. Unlike \RNAensign, it does not require a seed sequence to initiate its search. Finally, \ourprog implements a novel algorithm using weighted sampling techniques~\cite{Bodini2010} that enables us to control, for the first time, \textit{explicitly} the \GCContent of the solution. This functionality is essential because wild-type sequences within living organisms often present medium or low \GCContent, presumably to offer better transcription rates and/or structural plasticity. Previous programs do not allow to control this parameter and tend to output sequences with high \GCContent{}s.

### 2013-03-03 ISMB/main_ISMB.tex
typos

\begin{document}
\maketitle
%\section{}
%\subsection{}
\begin{abstract}
\noindent
\textbf{Motivations:} The design of RNA sequences folding into predefined secondary structures is a milestone for many synthetic biology and gene therapy studies. Most of the current software use similar local search strategies (i.e. a random seed is progressively adapted to acquire the desired folding propertues) and more importantly do not allow the user to control explicitly the nucleotide distribution such as the \GCContent in their sequences. However, the latter is an important criteria for large-scale applications as it could presumably be used to design sequences with better transcription rates and/or structural plasticity.\\
\noindent
\textbf{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \GCContent. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches  for sampling sequences with a specific GC content and target structure.\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

>>>>

\begin{document}
\maketitle
%\section{}
%\subsection{}
\begin{abstract}
\noindent
\textbf{Motivations:} The design of RNA sequences folding into predefined secondary structures is a milestone for many synthetic biology and gene therapy studies. Most of the current software uses similar local search strategies (i.e. a random seed is progressively adapted to acquire the desired folding propertues) and more importantly do not allow the user to control explicitly the nucleotide distribution such as the \GCContent in their sequences. However, the latter is an important criterion for large-scale applications as it could presumably be used to design sequences with better transcription rates and/or structural plasticity.\\
\noindent
\textbf{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \GCContent. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches  for sampling sequences with a specific GC content and target structure.\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

### 2013-03-03 ISMB/methods_ISMB.tex
typos

We introduce a probabilistic model for the design of RNA sequences with a specific \GCContent and folding into a predefined secondary structure.
For the sake of simplicity, we choose to base this proof-of-concept implementation on a simplified free-energy function $\ES(\cdot)$, which only considers the contributions of 
stacked canonical base-pairs. We show how a modification of the dynamic programming scheme used in the \RNAmutants allows for the sampling of good and diverse design candidates, in linear time and space complexities.

@@

Despite its apparent quadratic complexity, the recurrence described in Subsection~\ref{sec:pf} can be computed in linear $\Theta(|\Target|)$ time and space, owing to the fact that, on any recursive call, the position $j$ is entirely determined by $i$.

@@

\subsection{Postprocessing unpaired regions: A local/global (glocal) hybrid approach}
\label{subsec:glocal_method}
Due to our simplified our energy model, unpaired regions are not subject to design constraints other than the \GCContent, leading to modest probabilities for refolded design candidates to match the targeted structure. To improve these performances and test the complementarity  of our global sampling approach with previous contributions based on local search, we used the \RNAinverse software to redesign unpaired regions. We specified a constraint mask to prevent stacking base-pairs from being modified and, whenever necessary, reetablished their content content a posteriori as \RNAinverse has been witnessed to take some liberties with constraints masks. As shown in Table~\ref{table:impact_on_gc} (Supplementary material), this postprocessing does not drastically alter the \GCContent, so the glocal approach reasonably addresses the contrained \GCContent design problem.

>>>>

We introduce a probabilistic model for the design of RNA sequences with a specific \GCContent and folding into a predefined secondary structure.
For the sake of simplicity, we choose to base this proof-of-concept implementation on a simplified free-energy function $\ES(\cdot)$, which only considers the contributions of 
stacked canonical base-pairs. We show how a modification of the dynamic programming scheme used in \RNAmutants allows for the sampling of good and diverse design candidates, in linear time and space complexities.

@@

Despite its apparent quadratic complexity, the recurrence described in Subsection~\ref{sec:pf} can be computed in linear $\Theta(|\Target|)$ time and space, owing to the fact that, on any recursive call, the position $k$ is entirely determined by $i$.

@@

\subsection{Postprocessing unpaired regions: A local/global (glocal) hybrid approach}
\label{subsec:glocal_method}
Due to our simplified energy model, unpaired regions are not subject to design constraints other than the \GCContent, leading to modest probabilities for refolded design candidates to match the targeted structure. To improve these performances and test the complementarity  of our global sampling approach with previous contributions based on local search, we used the \RNAinverse software to redesign unpaired regions. We specified a constraint mask to prevent stacking base-pairs from being modified and, whenever necessary, reestablished their content a posteriori as \RNAinverse has been witnessed to take some liberties with constraints masks. As shown in Table~\ref{table:impact_on_gc} (Supplementary material), this postprocessing does not drastically alter the \GCContent, so the glocal approach reasonably addresses the contrained \GCContent design problem.

### 2013-03-03 ISMB/results_ISMB.tex
fix typo
typos

 To ease the visualization of the results, we clustered together strucures
 having similar length, stacks density and proportion of free-nucleotides in loops. 
  We present in Figure~\ref{fig:bins} the distribution of structures in each category.

@@

\subsection{Design}
 To benchmark our method, we first started by sampling about $100$ sequences per structure. After correcting the free nucleotides with
 \texttt{RNAinverse} we computed their MFE with \texttt{RNAfold} from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \ourprog in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \ourprog sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is, in general, not a good predictor for the hardness of designing a structure. 
Instead, a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solutions was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

@@

Also, a critical properties in RNA sequence design is 
the frequency of the MFE. 
The sequence should be in its target conformation long enough to
perform the desired action. Presented in Figure~\ref{fig:freq} we see that
there is a slow decline of the frequency with the increase in size. Yet,
for an average \GCContent, the frequency stays over $10\%$ even
at size of $100$ nucleotides.

>>>>

 To ease the visualization of the results, we clustered together strucures
 having similar length, stacks density and proportion of free nucleotides in loops. 
  We present in Figure~\ref{fig:bins} the distribution of structures in each category.

@@

\subsection{Design}
 To benchmark our method, we first started by sampling about $1000$ sequences per structure. After correcting the free nucleotides with
 \texttt{RNAinverse} we computed their MFE with \texttt{RNAfold} from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \ourprog in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \ourprog sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is, in general, not a good predictor for the hardness of designing a structure. 
Instead, a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solution was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

@@

Also, a critical propertie in RNA sequence design is 
the frequency of the MFE. 
The sequence should be in its target conformation long enough to
perform the desired action. Presented in Figure~\ref{fig:freq} we see that
there is a slow decline of the frequency with the increase in size. Yet,
for an average \GCContent, the frequency stays over $10\%$ even
at size of $100$ nucleotides.

### 2013-03-03 ISMB/supplementary_ISMB.tex
typos

%!TEX root = main_ISMB.tex
\section{Supplementary data}
\subsection{Benchmark \ourprog vs \RNAinverse}
Using the same dataset of $50$ structures, we generated $100$ samples
per structure with \RNAinverse. They yield for most parameters
a high entropy.
Table.~\ref{tab:nb_rnainv} contains the number of sample with a given \GCContent generated by \RNAinverse. The vast majority resides around $50\%-70\%$, showing one of its major deficiencies. 
 For all structures that have been solved 
by the three methods, only \RNAinverse, only \texttt{Incarnation} and
\texttt{Incarnation} followed by \RNAinverse, 
we present for every different concentration of \GCContent
the average sample entropy. and the average base pairs entropy in Figure.~\ref{fig:rnainverse}.

>>>>

%!TEX root = main_ISMB.tex
\section{Supplementary data}
\subsection{Benchmark \ourprog vs \RNAinverse}
Using the same dataset of $50$ structures, we generated $100$ samples
per structure with \RNAinverse. They yield for most parameters
a high entropy.
Table.~\ref{tab:nb_rnainv} contains the number of samples with a given \GCContent generated by \RNAinverse. The vast majority resides around $50\%-70\%$, showing one of its major deficiencies. 
 For all structures that have been solved 
by the three methods, only \RNAinverse, only \texttt{Incarnation} and
\texttt{Incarnation} followed by \RNAinverse, 
we present for every different concentration of \GCContent
the average sample entropy and the average base pairs entropy in Figure.~\ref{fig:rnainverse}.
